Non-word repetition in Yélî Dnye

### 2020-05-14 manuscript.Rmd

```{r setup, include=FALSE}
#Sys.setlocale("LC_CTYPE", "en_US_POSIX.UTF-8")
#Sys.setlocale("LANG", "en_US_POSIX.UTF-8")
Sys.setlocale(category = "LC_ALL", locale = "en_SE.utf8")

@@

@@

```{r}
read.csv("prevlit.csv")->kids

@@

@ac ADD TABLE SUMMARIZING PREVIOUS WORK

- age differences
- length effects

@@

The child was donned a headset (**xx @mc please fill in** for most of the children, xx @ac fill in for the rest). For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth. A local informant sat next to the child, to  provide the instructions and, if needed, coach the child to make sure that they understood that the task was to repeat the items precisely without changing anything using the three practice items as well as real words. The experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headhphones.

@@

```{r}
# read.table("log2.txt",header=T,sep="\t")->x
# x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
# x[order(x$out),]->x
# x$cor.online=1
# for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
#   
# subset(x,rep=="1")->x
# aggregate(x$cor.online,by=list(x$child),mean)->onl
# colnames(onl)<-c("ID","score.online")
# 
# #print(onl)
# 
# aggregate(x$corr,by=list(x$child),mean)->ms
# colnames(ms)<-c("ID","score")
# merge(ms,onl,all=T)->ms
# merge(ms,demo,all=T)->scores
# scores$Age=as.numeric(as.character(scores$Age))
# scores$Age[is.na(scores$Age)]<-15
# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
# 
# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
# abline(lm(scores$score*100~scores$Age))
# plot(scores$score~scores$monolingual)
# points(scores$score~as.numeric(scores$monolingual),pch=20)
# 
# plot(ms$score.online~ms$score)

read.csv("NWR-transcription.csv",sep=";")->coding

#remove some weird items
coding[coding$item!="task4",]->coding

read.csv("final_order_ful.txt",sep="\t")->coding_key
coding_key$row=row.names(coding_key)

merge(coding,coding_key,all=T)->joint

merge(joint,demo,by.x="id",by.y="ID",all.x=T)->joint

#order so we have all the items by a given child in the right chrono order
joint[order(joint$file, joint$int),]->joint

#add variable coding whether it's the child's first or subsequent attempts
joint$previous<-c(NA,as.character(joint$target[1:(dim(joint)[1]-1)]))
joint$attempt<-ifelse(joint$target!=joint$previous,"first","subsequent")
joint$attempt[1]<-"first"

#remove practice items
practice=joint[(joint$target %in% c("wi","poni","nopimade")),]
joint=joint[!(joint$target %in% c("wi","poni","nopimade")),]

read.csv("stimuli.csv",header=T)->stims

# phonologize things
merge(joint,stims,by="target",all=T)->joint
joint$target_pron=as.character(joint$ortho)

joint$pron=as.character(joint$mispronunciation)

correspondances=matrix( #list correspondances always by pairs
  c(
    #nasal long to unichar
    ":ii","ì",
    ":ee","è",
    ":aa","à",
    ":êê","3",
    ":ââ","6",
    ":uu","ù",
    ":oo","Ò",
    #long to unichar
    "ii","ï",
    "ee","ë",
    "éé","Ë",
    "áá","Ä",
    "aa","ä",
    "îî","Ï",
    "êê","2",
    "ââ","5",
    "uu","ü",
    "óó","ö",
    "oo","Ö",
    #short nasal
    ":i","1",
    ":e","4",
    ":a","ã",
    ":ê","6",
    ":â","7",
    ":u","8",
    ":o","9",
    #consonants, veeeery dirty!
    "'n","ń",
    "tp",":",
    "dp","/",
    "kp","P",
    "ngm","G",
    "ńm","M",
    "ng","N",
    "mb","b",
    "nd","D",
    "nt","T",
    "nk","K",
    "dn","?",
    "kn","!",
    "vy","V",
    "mt","C",
    "md","J",
    "mg","X",
    "dm","S",
    "km","Z",
    "gh","H",
    "lv","L"
  ),#last item above should not have a comma
  ncol = 2,byrow = T)

for(i in 1:dim(correspondances)[1]) {
  joint$pron=gsub(correspondances[i,1],correspondances[i,2],joint$pron)
  joint$target_pron=gsub(correspondances[i,1],correspondances[i,2],joint$target_pron)
}

#deal with hesitations
joint$hesitation=F

@@

@@

aggregate(joint$correct*100,by=list(joint$id,joint$age.exact,joint$attempt),mean)->pwcm_att
colnames(pwcm_att)<-c("ID","Age","Attempt","Word")

@@

```{r exclude subsequent}
joint=joint[joint$attempt=="first",]
```

>>>>

```{r setup, include=FALSE}
REDOALL=FALSE #use true if you want to recalculate the final-data and other data sheets

#Sys.setlocale("LC_CTYPE", "en_US_POSIX.UTF-8")
#Sys.setlocale("LANG", "en_US_POSIX.UTF-8")
Sys.setlocale(category = "LC_ALL", locale = "en_SE.utf8")

@@

correspondances=matrix( #list correspondances always by pairs, orthography then phonology
  c(
    #nasal  to unichar
    ":ii","ì",
    ":ee","è",
    ":aa","à",
    ":êê","3",
    ":ââ","6",
    ":uu","ù",
    ":oo","Ò",
    #long to unichar
    "ii","ï",
    "ee","ë",
    "éé","Ë",
    "áá","Ä",
    "aa","ä",
    "îî","Ï",
    "êê","2",
    "ââ","5",
    "uu","ü",
    "óó","ö",
    "oo","Ö",
    #short nasal
    ":i","1",
    ":e","4",
    ":a","ã",
    ":ê","6",
    ":â","7",
    ":u","8",
    ":o","9",
    #consonants, veeeery dirty!
    "'n","ń",
    "tp",":",
    "dp","/",
    "kp","P",
    "ngm","G",
    "ńm","M",
    "ng","N",
    "mb","b",
    "nd","D",
    "nt","T",
    "nk","K",
    "dn","?",
    "kn","!",
    "vy","V",
    "mt","C",
    "md","J",
    "mg","X",
    "dm","S",
    "km","Z",
    "gh","H",
    "lv","L"
  ),#last item above should not have a comma
  ncol = 2,byrow = T)

```
## TODO

- create stimuli table that has complexity & length
- integrate demo data & stim table in final data
- add analysis on the basis of complexity & length
- add analysis on the basis of demo chars
- fix the back-transformation
- add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
- discussion

@@

```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words."}
read.csv("prevlit.csv")->pl
#duplicate Piazzalunga logging short and long items separately
pl$Score[pl$Study=="Piazzalunga 2018"]<-pl$Short[pl$Study=="Piazzalunga 2018"]
pl$Study<-as.character(pl$Study)
pl$Study[pl$Study=="Piazzalunga 2018"]<-"Piazzalunga 2018 1-2 syll"
dim(pl)[1]->beg
pl=rbind(pl,pl[pl$Study=="Piazzalunga 2018 1-2 syll",])
pl[(beg+1):dim(pl)[1],"Score"]<-pl[(beg+1):dim(pl)[1],"Long"]
pl[(beg+1):dim(pl)[1],"Study"]<-"Piazzalunga 2018 3-4 syll"
pl$Study=paste(pl$Language,pl$Study)
library(lattice)
xyplot(Score~Age,groups=Study,data=pl,type=c('l','p'),auto.key = T ,pch=20)

@@

@@

The child was donned a headset (**xx @mc please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder. For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth. A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything. An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.

@@

```{r wrangling,eval=REDOALL}
source("wrangling.R")

@@

```{r readin}
read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
#create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
firstp=subset(trnsc,rep==1 & !is.na(rep))
```

Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(firstp$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(firstp$correct)*100)`% when the offline coding is considered. On average, the phoneme-based normalized Levenshtein distance was `r round(mean(firstp$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.

We then inspected whether accuracy varied as a function of word length.

```{r figlength,eval=F}
#not working
aggregate(firstp$correct,list(firstp$tarlen,firstp$id),mean)
plot(firstp$correct~firstp$tarlen)

```
- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?

@@

aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$attempt),mean)->pwcm_att
colnames(pwcm_att)<-c("ID","Age","Attempt","Word")

@@

```{r exclude subsequent}
trnsc=trnsc[trnsc$attempt=="first",]
```
